---
title: 'Assignment 1: Problem I (IEC7413)'
author: "Sonia Sofia"
date: "2025-05-11"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-libraries-incomeUK}
# Note: install these in the console before running the script:
# install.packages("Ecdat")
# install.packages("dynlm")
# install.packages("lmtest")
# install.packages("tseries")

library(Ecdat)      # Contains the IncomeUK dataset
library(dynlm)      # For dynamic linear models
library(lmtest)     # For Ljung-Box test
library(tseries)    # For Jarque-Bera test
```

```{r inspect-transform-income-data}
data(IncomeUK)

# Extract the income variable
y <- IncomeUK[, "income"]

## First-difference to induce stationarity
dy <- diff(y)

## Summary Statistics
head(IncomeUK)
summary(IncomeUK)

# Set up the plotting area: 1 row, 2 columns
par(mfrow = c(1, 2))

# Plot original income
plot(y,
     main = "UK Disposable Income (1971–1985)",
     ylab = "Income (nominal £)",
     xlab = "Time",
     type = "l",
     col = "darkblue",
     lwd = 2)

# Plot differenced income
plot(dy,
     main = "First Difference of Income",
     ylab = "ΔIncome",
     xlab = "Time",
     type = "l",
     col = "darkred",
     lwd = 2)
```

The income data ranges from approximately £9,000 to £60,000. The average is about £30,600, but the median is only £27,200, indicating a right-skewed distribution with some high-income outliers.

The top 25% of earners (above £44,866) make nearly three times more than the bottom 25% (below £15,820), reflecting a wide income spread and a clear upward trend over time.

The left plot confirms this trend in the raw series, while the right plot shows that after differencing, the data becomes much more stable. This validates differencing as an appropriate transformation before time series modeling.

```{r fit-ar1-to-ar5-models}
# Fit AR(1) to AR(5) models on differenced income
# difference of income
# All models should use the same number of observations

models <- list(
  AR1 = dynlm(dy ~ L(dy, 1)),
  AR2 = dynlm(dy ~ L(dy, 1:2)),
  AR3 = dynlm(dy ~ L(dy, 1:3)),
  AR4 = dynlm(dy ~ L(dy, 1:4)),
  AR5 = dynlm(dy ~ L(dy, 1:5))
)
```

## Question 1: Estimating AR(p) Models
We estimated five autoregressive (AR) models using the first-differenced income data, ranging from **AR(1)** to **AR(5)**.
Each model includes lagged values of the dependent variable and is stored in a list, allowing for efficient comparison, AIC evaluation, and residual diagnostics.

```{r evaluate-aic-select-best-ar}
# We compare AR models using Akaike's Information Criterion 
# (AIC). 
# The goal is to select the model with the lowest AIC

AIC_values <- sapply(models, AIC)
print("AIC values for each AR model:")
print(AIC_values)

# Best model according to AIC:
best_model <- names(which.min(AIC_values))
cat("\nBest model according to AIC is:", best_model, "\n")
```
## Question 2: AIC Comparison
Out of the five AR models tested, **AR(5)** had the lowest AIC (828.04), with a 13-point improvement over **AR(4)**. Lower AIC values indicate better model fit, penalizing complexity.
According to Burnham and Anderson (2002), a difference greater than 10 in AIC provides **strong evidence** in favor of the model with the lower value. This means AR(5) is the most appropriate model in the set.

```{r diagnostic-tests-ar-models}
# Perform Ljung-Box and Jarque-Bera tests on model residuals
# Interpret and compare results with AIC conclusions
for (name in names(models)) {
  cat("\nModel:", name, "\n")
  residuals <- resid(models[[name]])
  lb_test <- Box.test(residuals, lag = 10, type = "Ljung-Box")
  jb_test <- jarque.bera.test(residuals)
  print(lb_test)
  print(jb_test)
}
print(round(AIC_values, 2))
```

## Question 3: Residual Diagnostics
We tested the AR models using two diagnostic tools: the **Ljung–Box test** (to check for autocorrelation) and the **Jarque–Bera test** (to check for normality of residuals).
We used **lag = 10** because our dataset has 58 observations, and √n ≈ 7.6. Ten lags cover approximately 2.5 years of quarterly data, which aligns with typical macroeconomic practice. 
**AR(4)** and **AR(5)** passed both tests, but since **AR(5)** also had the lowest AIC, it is the best overall model for inference and forecasting.

```{r summary-model-ar5-selection}
cat("\nSummary:\n")
cat("The AR(5) model had the lowest AIC and passed both residual tests,\n")
cat("indicating no autocorrelation and approximate normality.\n")
cat("This model offers the best overall balance of fit and residual quality,\n")
cat("making it suitable for interpretation and forecasting purposes.\n")
```

## Key Takeaway
AR(5) is both statistically justified and practically interpretable, making it a strong candidate for further forecasting applications.
---

[Click here to view the full HTML report](assignment1_problem1.html)
